---
title: "R Notebook"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
# Load necessary libraries
library(bigrquery)
library(dplyr)
library(ggplot2)
library(lubridate)
```

### Set Project and Dataset Variables

```{r}
# Define project and dataset variables
project_id <- "fastdate"
dataset_id <- "fastdate.new_york_citibike.merged_citibike_data"
```

### Query Data from BigQuery

```{r}
# SQL query to fetch data
query <- "
  SELECT
    tripduration,
    starttime,
    stoptime,
    start_station_name,
    end_station_name,
    bikeid,
    usertype,
    birth_year,
    gender,
    start_latitude,
    start_longitude,
    end_latitude,
    end_longitude
  FROM
    `fastdate.new_york_citibike.merged_citibike_data`
  WHERE
    tripduration IS NOT NULL
    AND start_latitude IS NOT NULL
    AND start_longitude IS NOT NULL
    AND end_latitude IS NOT NULL
    AND end_longitude IS NOT NULL
  LIMIT 1000000
"

# Execute query and download data
citibike_data <- bq_project_query(project_id, query) %>%
  bq_table_download()
```

### Data Cleaning and Transformation

```{r}
# Clean and transform data
citibike_data <- citibike_data %>%
  mutate(
    starttime = as.POSIXct(starttime, format = "%Y-%m-%d %H:%M:%S"),
    stoptime = as.POSIXct(stoptime, format = "%Y-%m-%d %H:%M:%S"),
    trip_duration_minutes = tripduration / 60,
    trip_duration_hours = tripduration / 3600, # Convert seconds to hours
    age = ifelse(!is.na(birth_year), year(Sys.Date()) - birth_year, NA)
  ) %>%
  filter(age >= 10 & age <= 90)  # Filter out extreme ages

```

### Summary Statistics

```{r}
# Calculate summary statistics
summary_stats <- citibike_data %>%
  summarize(
    avg_trip_duration = mean(tripduration),
    median_trip_duration = median(tripduration),
    max_trip_duration = max(tripduration),
    total_trips = n(),
    avg_age = mean(age, na.rm = TRUE)
  )

print(summary_stats)
```

### Top 10 Popular Start Stations

```{r}
# Identify top 10 popular start stations
popular_stations <- citibike_data %>%
  group_by(start_station_name) %>%
  summarize(trips = n()) %>%
  arrange(desc(trips)) %>%
  top_n(10, trips)

print(popular_stations)
```

### Analyze Usage Patterns by Time of Day

```{r}
# Analyze usage patterns by hour of the day
time_of_day_stats <- citibike_data %>%
  mutate(hour = hour(starttime)) %>%
  group_by(hour) %>%
  summarize(
    avg_trip_duration = mean(tripduration),
    total_trips = n()
  )
print(time_of_day_stats)
```

### Analyze Usage Patterns by Day of the Week

```{r}
# Analyze usage patterns by day of the week
day_of_week_stats <- citibike_data %>%
  mutate(weekday = weekdays(starttime)) %>%
  group_by(weekday) %>%
  summarize(
    avg_trip_duration = mean(tripduration),
    total_trips = n()
  )
print(day_of_week_stats)
```

### Gender and Age Analysis

```{r}
# Analyze age and trip counts by gender
gender_age_stats <- citibike_data %>%
  group_by(gender) %>%
  summarize(
    avg_age = mean(age, na.rm = TRUE),
    total_trips = n()
  )
print(gender_age_stats)
```


### Visualizations

#### Geographic Distribution of Start Stations

```{r}
# Visualize geographic distribution of start stations
ggplot(citibike_data, aes(x = start_longitude, y = start_latitude)) +
  geom_point(alpha = 0.5, color = "blue") +
  labs(
    title = "Geographic Distribution of Start Stations",
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_minimal()


```


#### Top 10 Popular Start Stations

```{r}
ggplot(popular_stations, aes(x = reorder(start_station_name, -trips), y = trips)) +
  geom_bar(stat = "identity", fill = "blue") +
  coord_flip() +
  labs(title = "Top 10 Popular Start Stations", x = "Station Name", y = "Number of Trips") +
  theme_minimal()
```



#### Usage Patterns by Time of Day

```{r}
ggplot(time_of_day_stats, aes(x = hour, y = total_trips)) +
  geom_line(color = "red") +
  labs(title = "Usage Patterns by Time of Day", x = "Hour of the Day", y = "Number of Trips") +
  theme_minimal()
```
### Trip Duration Distribution
```{r}
# Distribution of trip duration
trip_duration_distribution <- citibike_data %>%
  filter(trip_duration_minutes <= 60)  # Exclude extreme values for visualization

ggplot(trip_duration_distribution, aes(x = trip_duration_minutes)) +
  geom_histogram(binwidth = 5, fill = "skyblue", color = "black") +
  labs(
    title = "Distribution of Trip Durations",
    x = "Trip Duration (Minutes)",
    y = "Frequency"
  ) +
  theme_minimal()

```


#### Usage Patterns by Day of the Week

```{r}
ggplot(day_of_week_stats, aes(x = reorder(weekday, total_trips), y = total_trips)) +
  geom_bar(stat = "identity", fill = "purple") +
  labs(title = "Usage Patterns by Day of the Week", x = "Day", y = "Number of Trips") +
  theme_minimal()
```
#### Age Distribution of Riders

```{r}
# Age distribution analysis
age_distribution <- citibike_data %>%
  filter(!is.na(age)) %>%
  group_by(age) %>%
  summarize(total_trips = n())

# Visualization: Age Distribution
ggplot(age_distribution, aes(x = age, y = total_trips)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(title = "Age Distribution of Riders", x = "Age", y = "Number of Trips") +
  theme_minimal()

```

#### Average Age by Gender

```{r}
ggplot(gender_age_stats, aes(x = gender, y = avg_age, fill = gender)) +
  geom_bar(stat = "identity") +
  labs(title = "Average Age by Gender", x = "Gender", y = "Average Age") +
  theme_minimal()
```

#### User Type Breakdown

```{r}
# Analyze user type breakdown
user_type_stats <- citibike_data %>%
  group_by(usertype) %>%
  summarize(total_trips = n())

ggplot(user_type_stats, aes(x = usertype, y = total_trips, fill = usertype)) +
  geom_bar(stat = "identity") +
  labs(
    title = "User Type Breakdown",
    x = "User Type",
    y = "Number of Trips"
  ) +
  theme_minimal()

```


### Clean Up

```{r}
# Free up memory
rm(citibike_data)
gc()
```
